White Dwarf Mergers on Adaptive Meshes\\ II. Collisions

### 2016-02-14 papers/wdmerger_II/paper.tex
Add discussion of the dual energy switch to paper II

Finally, we return to a point we hinted at above. The timestep
will only actually satisfy the energy criterion
$\Delta t \leq f_e \tau_e$ and species criterion
$\Delta t \leq f_{X_k} \tau_{X_k}$ when the estimates for
$\dot{e}$ and $\dot{X_k}$ we generate are at least as large
as the actual rate of change of energy and mass fractions
over the timestep. However, this can assumption can fail
quite radically during periods of runaway burning when
the rate of change of these quantities is highly nonlinear.
We do not want to neglect the errors caused by this approximation
because they may build up over an extended period of nonlinear
evolution and perhaps substantially change the final results.
To this end, we have implemented a timestep retry option in
\castro. The retry option works by first saving a copy of the
full state, and then doing the advance. We then explicitly
check whether the timestep criteria were violated during the advance,
using the time-centered energy and mass fractions and a
time-centered differencing of the old and new states to
calculate the evolution timescale. If so, we reject this
advance and restore the original state data. We then perform
a subcycled update, with the maximum timestep chosen so that
the criterion is satisfied in each subcycle. Note that this
occurs entirely on a single level and is separate from the
subcycling that is used by \boxlib\ to advance the different
AMR levels in time; the full AMR algorithm does not ever
know that the retry subcycling took place. This can help
substantially to mollify the problem described above, where
the time derivative estimation methods proposed were not
sufficient to obtain the desired limiting. Thus we strongly
recommend that hydrodynamics codes implement such an approach,
which ensures that the coupling actually does stay at the
level requested. The most significant drawback is that saving
a copy of the old state effectively doubles the memory footprint
of the code, so this option should be disabled when not needed.
However, it may be crucial in determining the evolution of
systems such as white dwarf collisions, where even a short
period of incorrect advances can significantly pollute the
final answer, and the additional memory expense is justified
by obtaining a more accurate solution.

@@

\clearpage
\appendix

>>>>

Now we return to a point we hinted at above. The timestep
will only actually satisfy the energy criterion
$\Delta t \leq f_e \tau_e$ and species criterion
$\Delta t \leq f_{X_k} \tau_{X_k}$ when the estimates for
$\dot{e}$ and $\dot{X_k}$ we generate are at least as large
as the actual rate of change of energy and mass fractions
over the timestep. However, this can assumption can fail
quite radically during periods of runaway burning when
the rate of change of these quantities is highly nonlinear.
We do not want to neglect the errors caused by this approximation
because they may build up over an extended period of nonlinear
evolution and perhaps substantially change the final results.
To this end, we have implemented a timestep retry option in
\castro. The retry option works by first saving a copy of the
full state, and then doing the advance. We then explicitly
check whether the timestep criteria were violated during the advance,
using the time-centered energy and mass fractions and a
time-centered differencing of the old and new states to
calculate the evolution timescale. If so, we reject this
advance and restore the original state data. We then perform
a subcycled update, with the maximum timestep chosen so that
the criterion is satisfied in each subcycle. Note that this
occurs entirely on a single level and is separate from the
subcycling that is used by \boxlib\ to advance the different
AMR levels in time; the full AMR algorithm does not ever
know that the retry subcycling took place. This can help
substantially to mollify the problem described above, where
the time derivative estimation methods proposed were not
sufficient to obtain the desired limiting. Thus we strongly
recommend that hydrodynamics codes implement such an approach,
which ensures that the coupling actually does stay at the
level requested. The most significant drawback is that saving
a copy of the old state effectively doubles the memory footprint
of the code, so this option should be disabled when not needed.
However, it may be crucial in determining the evolution of
systems such as white dwarf collisions, where even a short
period of incorrect advances can significantly pollute the
final answer, and the additional memory expense is justified
by obtaining a more accurate solution.

One other point worth noting for the coupling of the reactions
and the hydrodynamics relates to the dual-energy formalism used
by \castro\ and many other hydrodynamics codes \cite{bryan:1995}.
\castro\ evolves separately equations for the total energy $(\rho E)$ and
for the internal energy $(\rho e)$. The former is conservative while the latter
is not, so for accurate hydrodynamics we prefer to use the total
energy when possible and to use the evolved internal energy variable
only in situations where the kinetic energy dominates the contribution
to the total energy. Indeed, for the cosmological purpose for which
this was originally developed, \cite{ENZO} choose a value $\eta_2 = 0.1$
so that $e$ is only updated to be equal to $E - \mathbf{v}^2/2$ if
$e > \eta_2\, E$. In other cases the evolved value of the internal
energy variable is preserved. However, this choice does not work
for our application, because the ultimate cause of the nuclear
burning in a white dwarf collision is the rapid conversion of
kinetic energy from the white dwarf bulk motion to thermal energy
as the white dwarfs slam into each other. Keeping $\eta_2$ large
prevents this from happening, and consequently the temperature
will never reach values high enough to generate significant
amounts of nickel. For this paper we choose $\eta_2 = 10^{-4}$,
which is low enough for the correct conversion of kinetic energy
but not so low that we need to be concerned about roundoff issues
caused by subtracting kinetic from total energy. This is also
the value that was used by \cite{hawley:2012}.

@@
